A numerical study of fixation probabilities for strategies in the Iterated Prisoner's Dilemma

### 2017-04-14 tex/main.tex
Redraw moran process diagram.
Use hashing to avoid counting on color.
Make minor stylistic tex changes.
Remove MH comment no longer relevant because N/2.

\usepackage{tikz}
\usetikzlibrary{calc, shapes}

@@

Figure~\ref{fig:moran_process} shows a diagrammatic representation of the Moran
process. The Moran process is a stochastic birth death process on a finite
population in which the population size stays constant over time. Individuals
are \textbf{selected} according to a given fitness landscape. Once selected, a
given individual is reproduced and similarly another individual is chosen to be
removed from the population. In some settings mutation is also considered but
without mutation (the case considered in this work) this process will arrive at
an absorbing state where the population is entirely made up of a single
individual. The probability with which a given strategy is the survivor is call
the absorption probability. A more detailed analytic description of this is
given in Section~\ref{sec:validation}.

\begin{figure}[!hbtp]
    \centering
    \input{../img/moran_process.tex}
    \label{fig:moran_process}
    \caption{A diagrammatic representation of a Moran process}
\end{figure}

@@

\begin{itemize}
    \item The Axelrod library (IPD strategies and Moran processes)
        \cite{axelrodproject}.
    \item The matplotlib library (visualisation) \cite{hunter2007matplotlib}.
    \item The pandas and numpy libraries (data manipulation)
        \cite{mckinney2010data, walt2011numpy}
\end{itemize}

@@

\begin{equation}\label{equ:fixation_probability}
x_i = \frac{1 + \sum_{j=1}^{i-1}\prod_{k=1}^{j}\gamma_j}{1 + \sum_{j=1}^{N-1}\prod_{k=1}^{j}\gamma_j}
\end{equation}

@@

Figure~\ref{fig:comparison_stochastic} shows the fixation probabilities for
stochastic strategies. These are no longer a good match which highlights the
weakness of the analytical formulae the relies on the average payoffs. A
detailed analysis of the 172 strategies considered will be shown in the next
Section.

@@

\[ \rho = \frac{1 - r^{-i}}{1 - r^{-N}}\]
and $\rho = 1 / N$ if $r=1$ (the neutral fixation probability).

This corresponds to a game matrix [[1, 1], [r, r]] (or [[r, r], [1, 1]]), which
is of course not what we have -- it's a little complicated because our "fitness"
is not the payout from the game matrix, rather the sum of the total scores of
all the interactions each round. So ALLC and TFT are neutral wrt to each other
because they will have the same score each round, giving an effective fitness
landscape $f(i, N-i) = A [i, N-i]^T$ given by the matrix $A = [[1,1],[1,1]]$.
This means that noise and the number of turns per Moran round are significant
parameters. I think we should fix the turns at 200; some recent authors run the
turns to infinity (to reach stationarity on the sub-"Markov process" on the
states (C, C), (C, D), (D, C), (D, D)) but we can't analytically compute the
stationary distribution for strategies that use more than one round of memory
(and it's not really a Markov process for more than one round of memory anyway).
Plus it's unrealistic, and ultimately just amounts to a transform of the game
matrix.

@@

A nice addition would be, for an interesting combination of strategies, to
measure the fixation value for all $(i, N-i)$ and compare to the above formula
for the value of $r$ derived from the $(1, N-1)$ case. This would show how much
we deviate from frequency independence.

The existing notebook attempts to get at 1 and 2 by looking at the distributions
of fixation probabilities for each strategy -- that's what the box plots for
each N try to visualize for particular N, and the "Player Rankings by Median vs.
Population Size" for how the cooperative strategies become more successful as N
increases. That plot is the main takeaway IMO, and reinforces the "evolution of
cooperation" narrative that's so popular. We can tie back to Press and Dyson
here -- yes, ZD strategies are good Head-to-Head and in small populations, but
they aren't great when the population size gets bigger. How much bigger? Even at
N=4 there is a dramatic decline for ZD-extort. Note that this goes against the
claims of Stewart and Plotkin (they claimed that ZD strategies basically
dominate the Moran process no matter how much memory you allow). This also
matches our tournament results -- ZD strategies win matches but not tournaments.

@@

Beyond the raw data, we should try to estimate the strategies that are
1) most resistant to invasion
2) the best invaders
3) "most neutral"

as a function of N across the entire population of strategies. This can really
open up if you want to say optimize a parameterized strategy to be most
resistant to invasion (a topic of future work, perhaps) -- for example Random(p)
for what p is best?

>>>>

\usepackage{tikz}
\usetikzlibrary{calc, shapes, patterns}

@@

Figure~\ref{fig:moran_process} shows a diagrammatic representation of the Moran
process. The Moran process is a stochastic birth death process on a finite
population in which the population size stays constant over time. Individuals
are \textbf{selected} according to a given fitness landscape. Once selected, a
given individual is reproduced and similarly another individual is chosen to be
removed from the population. In some settings mutation is also considered but
without mutation (the case considered in this work) this process will arrive at
an absorbing state where the population is entirely made up of a single
individual. The probability with which a given strategy is the survivor is
called the absorption probability. A more detailed analytic description of this
is given in Section~\ref{sec:validation}.

\begin{figure}[!hbtp]
    \centering
    \input{../img/moran_process.tex}
    \caption{A diagrammatic representation of a Moran process}
    \label{fig:moran_process}
\end{figure}

@@

\begin{itemize}
    \item The Axelrod library (IPD strategies and Moran processes)
        \cite{axelrodproject}.
    \item The matplotlib library (visualisation) \cite{hunter2007matplotlib}.
    \item The pandas and numpy libraries (data manipulation)
        \cite{mckinney2010data, walt2011numpy}.
\end{itemize}

@@

\begin{equation}\label{equ:fixation_probability}
x_i = \frac{1 + \sum_{j=1}^{i-1}\prod_{k=1}^{j}\gamma_j}{1 + \sum_{j=1}^{N-1}
      \prod_{k=1}^{j}\gamma_j}
\end{equation}

@@

Figure~\ref{fig:comparison_stochastic} shows the fixation probabilities for
stochastic strategies. These are no longer a good match which highlights the
weakness of the analytical formulae that relies on the average payoffs. A
detailed analysis of the 172 strategies considered will be shown in the next
Section.

@@

@@

A nice addition would be, for an interesting combination of strategies, to
measure the fixation value for all $(i, N-i)$ and compare to the above formula
for the value of $r$ derived from the $(1, N-1)$ case. This would show how much
we deviate from frequency independence.
The existing notebook attempts to get at 1 and 2 by looking at the distributions
of fixation probabilities for each strategy -- that's what the box plots for
each N try to visualize for particular N, and the "Player Rankings by Median vs.
Population Size" for how the cooperative strategies become more successful as N
increases. That plot is the main takeaway IMO, and reinforces the "evolution of
cooperation" narrative that's so popular. We can tie back to Press and Dyson
here -- yes, ZD strategies are good Head-to-Head and in small populations, but
they aren't great when the population size gets bigger. How much bigger? Even at
N=4 there is a dramatic decline for ZD-extort. Note that this goes against the
claims of Stewart and Plotkin (they claimed that ZD strategies basically
dominate the Moran process no matter how much memory you allow). This also
matches our tournament results -- ZD strategies win matches but not tournaments.

@@
